# Heart Failure Prediction Dataset Analysis

## Packages

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(gtsummary)
library(gt)
library(corrplot)
library(broom)
library(webshot2)
```

## Importing the file

```{r}
heart <- read_csv("heart.csv")
```

## Pre-Processing

Serum Cholesterol values equal to 0 are clinically imposible, and as they are included in the dataset, they were removed before the statistical analysis.

```{r}
heart=heart[heart$Cholesterol!=0,]
```

## Univariate Analysis

### Sample Size

```{r}
size=nrow(heart)
```

### Sample Distribution by Age

```{r}
ggplot(heart,
       aes(x = cut(
         Age,
         breaks = seq(20, 80, by = 5),
         right  = FALSE,
         labels = paste(
           seq(20, 75, by = 5),
           seq(24, 79, by = 5),
           sep = "–"
         )
       ))) +
  geom_bar(fill="grey70",color="grey55") +
  theme_minimal() +
  labs(x = "Age Range (years)",
       y = "Number of Patients")
```

### Sample Distribution by Sex

```{r}
sex_props=prop.table(table(heart$Sex))
pie(table(heart$Sex),
    labels=paste0(c("Female","Male"), " (", round(100*sex_props), "%)"),
    col=c("pink","lightblue"),
    border="white",
    main="Sample Distribution by Sex",
    radius=1,
    cex=1.2)
legend("right",
       legend = c("Female", "Male"),
       fill   = c("pink", "lightblue"),
       bty    = "n",
       cex=1.1)
mtext(
  side = 1, line = 3,
  text = paste0(
    "Sample Size: ", length(heart$Sex),
    " (Female: ", table(heart$Sex)["F"], ", Male: ", table(heart$Sex)["M"], ")"
  ),
  cex = 0.9,
  bty="n",
)
```

### Incidence of each type of Chest Pain

```{r}
ggplot(heart, aes(x = ChestPainType, fill = Sex)) +
  geom_bar(color = "black") + 
  scale_fill_manual(values = c("F" = "pink", "M" = "lightblue")) +
  labs(
    x = "Chest Pain Type",
    y = "Number of Occurrences",
    fill = "Sex",
    caption = "ASY = Asymptomatic   ATA = Atypical Angina   NAP = Non-anginal Pain   TA = Typical Angina"
  ) +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0.5)  # center at bottom
  )+
  ggtitle("Incidence of each type of Chest Pain",)+
    theme(
    plot.caption = element_text(hjust = 0.5),
    plot.title   = element_text(hjust = 0.5)
  )
```

### Summary Table

```{r}
tab1 = heart %>%
  select(Age, RestingBP, Cholesterol, MaxHR, Oldpeak,
         Sex, ChestPainType, RestingECG, ExerciseAngina, 
         ST_Slope, FastingBS, HeartDisease) %>%
  tbl_summary(
    type=all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c(
        "Median (IQR)" = "{median} ({p25}, {p75})", 
        "Range"        = "({min}, {max})"
      ),
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 1,
      all_categorical() ~ c(0, 0)
    ),
    label = c(
      Age ~ "Age (years)",
      RestingBP ~ "Resting Blood Pressure (mm Hg)",
      Cholesterol ~ "Serum Cholesterol (mm/dl)",
      MaxHR ~ "Maximum Heart Rate (bpm)",
      Oldpeak ~ "ST Depression (Oldpeak)",
      Sex ~ "Sex",
      ChestPainType ~ "Chest Pain Type",
      RestingECG ~ "Resting ECG",
      ExerciseAngina ~ "Exercise Induced Angina",
      ST_Slope ~ "ST Segment Slope",
      FastingBS ~ "Fasting Blood Sugar > 120 mg/dl",
      HeartDisease ~ "Heart Disease"
    )
  ) %>%
  modify_caption("**Table 1. Characteristics of Sample**") %>%
  modify_header(label = "**Variable**")
tab1

# Para salvar como imagem
# gt_tab <- as_gt(tab1)
# gtsave(gt_tab, "Table1.png")
```

## Multivariate Analysis

### Checking Normality

To check normality on quantitative variables (Age, RestingBP, Cholesterol, MaxHR and Oldpeak), we used the Shapiro-Wilk normality test. This is useful to choose between parametric or nonparametric tests.

```{r}
continuous_vars <- heart %>% 
  select(Age, RestingBP, Cholesterol, MaxHR, Oldpeak)
normality_tests <- lapply(continuous_vars, shapiro.test)

sapply(normality_tests, function(x) x$p.value)

#No variable has a normal distribution
```

### Age Distribution by Heart Disease

```{r}
heart_plot <- heart %>%
  mutate(
    HeartDisease = factor(HeartDisease, levels = c(0, 1), labels = c("No", "Yes"))
  )
ggplot(heart_plot, aes(x = HeartDisease, y = Age, fill = HeartDisease)) +
  geom_boxplot() + 
  scale_fill_manual(
    values = c("No" = "grey70", "Yes" = "red2") 
  ) +
  labs(
    title = "Age Distribution by Heart Disease Status",
    x = "Heart Disease Status",
    y = "Age (Years)",
    fill = "Heart Disease"
  ) +
  theme_minimal()+
  theme(legend.position = "none",
        plot.title   = element_text(hjust = 0.5)) 

#Wilcoxon Test
wilcox.test(Age ~ HeartDisease, data=heart)
#p-value < 0.05, H0 is rejected, Age is significantly different between groups with and without heart disease
```

### Heart Disease by Sex

```{r}
heart_plot <- heart %>%
  mutate(
    HeartDisease = factor(HeartDisease, levels = c(0, 1), labels = c("No", "Yes")), Sex=factor(Sex, levels=c("F","M"),labels=c("Female","Male"))
  )
ggplot(heart_plot, aes(x = Sex, fill = HeartDisease)) +
  
  geom_bar(color="black",position = "fill") +
  scale_fill_manual(values=c("No"="grey70","Yes"="red2"))+
  labs(
    title = "Proportion of Heart Disease by Sex",
    x = "Sex",
    y = "Proportion",
    fill = "Heart Disease") +
  theme_minimal() +
  theme(plot.title   = element_text(hjust = 0.5))+
  scale_y_continuous(labels = scales::percent)

#Chi-Squared Test
chisq.test(table(heart$Sex,heart$HeartDisease))
#p-value < 0.05, H0 is rejected, Heart Disease is significantly different between sexes
```

### Heart Disease by Type of Chest Pain

```{r}
heart_plot <- heart %>%
  mutate(
    HeartDisease = factor(HeartDisease, levels = c(0, 1), labels = c("No", "Yes")),
    ChestPainType = factor(ChestPainType)
  )
ggplot(heart_plot, aes(x = ChestPainType, fill = HeartDisease)) +
  
  geom_bar(color="black",position = "fill") +
  scale_fill_manual(values=c("No"="grey70","Yes"="red2"))+
  labs(
    title = "Proportion of Heart Disease by Type of Chest Pain",
    x = "Type of Chest Pain",
    y = "Proportion",
    fill = "Heart Disease",
    caption = "ASY = Asymptomatic   ATA = Atypical Angina   NAP = Non-anginal Pain   TA = Typical Angina") +
  theme_minimal() +
  theme(plot.title   = element_text(hjust = 0.5))+
  scale_y_continuous(labels = scales::percent)
```

### Resting Blood Pressure by Heart Disease

```{r}
heart_plot <- heart %>%
  mutate(
    HeartDisease = factor(HeartDisease, levels = c(0, 1), labels = c("No", "Yes"))
  )
par(mfrow=c(1, 2)) 
ggplot(heart_plot, aes(x = HeartDisease, y = RestingBP, fill = HeartDisease)) +
  geom_boxplot() + 
  scale_fill_manual(
    values = c("No" = "grey70", "Yes" = "red2") 
  ) +
  labs(
    title = "Resting BP by Heart Disease",
    x = "Heart Disease Status",
    y = "Resting BP (mm Hg)",
    fill = "Heart Disease"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
wilcox.test(RestingBP ~ HeartDisease, data=heart)
```

### Cholesterol by Heart Disease

```{r}
ggplot(heart_plot, aes(x = HeartDisease, y = Cholesterol, fill = HeartDisease)) +
  geom_boxplot() + 
  scale_fill_manual(
    values = c("No" = "grey70", "Yes" = "red2") 
  ) +
  labs(
    title = "Cholesterol by Heart Disease",
    x = "Heart Disease Status",
    y = "Cholesterol (mm/dl)",
    fill = "Heart Disease"
  ) +
  theme_minimal() +
  theme(legend.position = "none", plot.title   = element_text(hjust = 0.5))
par(mfrow=c(1, 1))

wilcox.test(Cholesterol ~ HeartDisease, data=heart)
```

### Max Heart Rate by Age

```{r}
heart_scatter <- heart %>%
  mutate(HeartDisease = factor(HeartDisease, labels = c("No", "Yes")))
scatter_plot_multivariavel <- ggplot(heart_scatter, aes(x = Age, y = MaxHR, color = HeartDisease)) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("No" = "grey65", "Yes" = "red2")) +
  labs(
    title = "Max Heart Rate by Age",
    x = "Age (years)",
    y = "Max Heart Rate (BPM)",
    color = "Heart Disease"
  ) +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
scatter_plot_multivariavel
```

### Heart Disease by Exercise Angina

```{r}
heart_plot <- heart %>%
  mutate(
    HeartDisease = factor(HeartDisease, levels = c(0, 1), labels = c("No", "Yes")),
    ExerciseAngina = factor(ExerciseAngina, levels = c(0, 1), labels = c("No", "Yes")),
    ST_Slope = factor(ST_Slope)
  )
ggplot(heart_plot, aes(x = ExerciseAngina, fill = HeartDisease)) +
  geom_bar(color="black",position="fill") +
  scale_fill_manual(values = c("No" = "grey70", "Yes" = "red2")) +
  labs(
    title = "Heart Disease Proportion by Exercise Angina",
    x = "Exercise Angina",
    y = "Proportion",
    fill = "Heart Disease"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent)

chisq.test(table(heart$ExerciseAngina, heart$HeartDisease))
```

### Heart Disease by ST Slope

```{r}
ggplot(heart_plot, aes(x = ST_Slope, fill = HeartDisease)) +
  geom_bar(color="black", position = "fill") +
  scale_fill_manual(values = c("No" = "grey70", "Yes" = "red2")) +
  labs(
    title = "Heart Disease Proportion by ST Slope",
    x = "ST Slope (Peak Exercise)",
    y = "Proportion",
    fill = "Heart Disease"
  ) +
  theme_minimal() + theme(plot.title   = element_text(hjust = 0.5))+
  scale_y_continuous(labels = scales::percent)

chisq.test(table(heart$ST_Slope, heart$HeartDisease))
```

### Summary Table by Heart Disease

```{r}
heart_bivariavel <- heart %>%
  mutate(HeartDisease = factor(HeartDisease, 
                               levels = c(0, 1), 
                               labels = c("No", "Yes"))) %>%

  select(Age, RestingBP, Cholesterol, MaxHR, Oldpeak,
         Sex, ChestPainType, RestingECG, ExerciseAngina, 
         ST_Slope, FastingBS, HeartDisease) 
  
tabela_bivariada = heart_bivariavel %>%
  tbl_summary(
    by = HeartDisease,
    
    statistic = list(
      all_continuous() ~ c(
        "Median (IQR)" = "{median} ({p25}, {p75})", 
        "Range"        = "({min}, {max})"
      ),
      all_categorical() ~ "{n} ({p}%)"
    ),
    
    type = all_continuous() ~ "continuous2",
    digits = list(
      all_continuous() ~ 1,      
      all_categorical() ~ c(0, 1) 
    ),
    
    label = list(
      Age ~ "Age (years)",
      RestingBP ~ "Resting Blood Pressure (mm Hg)",
      Cholesterol ~ "Serum Cholesterol (mm/dl)",
      MaxHR ~ "Maximum Heart Rate (bpm)",
      Oldpeak ~ "ST Depression (Oldpeak)",
      Sex ~ "Sex",
      ChestPainType ~ "Chest Pain Type",
      RestingECG ~ "Resting ECG",
      ExerciseAngina ~ "Exercise Induced Angina",
      ST_Slope ~ "ST Segment Slope",
      FastingBS ~ "Fasting Blood Sugar > 120 mg/dl"
    )
  ) %>%
  
  add_p(
    test = list( 
        all_continuous() ~ "wilcox.test",
        all_categorical() ~ "chisq.test"  
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3) 
  ) %>%
  
  modify_header(label = "**Variable**", 
                stat_by = "**{level}**") %>% 
  modify_caption("**Table 2. Sample Characteristics by Heart Disease**") %>%
  modify_footnote(
    update = all_stat_cols() ~ "Quantitative Variables: Median (IQR), Wilcoxon Test | Categorical Variables: N (%), Chi-Squared Test."
  )

tabela_bivariada
```

### Heatmap

```{r}
heart_matrix_data <- heart %>%
  select(Age, RestingBP, Cholesterol, MaxHR, Oldpeak) 
matriz_cor_completa <- cor(heart_matrix_data,method=c("spearman"))
corrplot(
  matriz_cor_completa, 
  method = "color",                 
  type = "upper",                   
  tl.col = "black",                 
  tl.srt = 45,                      
  addCoef.col = "black",            
  number.cex = 0.7,                 
  title = "Heatmap", 
  mar = c(0, 0, 3, 0),
  diag=F,
)
```

## Model

```{r}
heart_final <- heart %>%
  mutate(ST_Slope = as.factor(ST_Slope)) %>% 
  mutate(ST_Slope = relevel(ST_Slope, ref = "Flat"))
modelo_logistico_final <- glm(
  HeartDisease ~ Age + Sex + ChestPainType + RestingBP + 
                 Cholesterol + FastingBS + RestingECG + 
                 MaxHR + ExerciseAngina + Oldpeak + ST_Slope, 
  data = heart_final, 
  family = binomial
)
tbl_modelo_logistico_final <- tbl_regression(
  modelo_logistico_final, 
  exponentiate = TRUE 
) %>%
  add_n() %>%
  modify_footnote(
    update = all_stat_cols() ~ paste0("Total N = ", nobs(modelo_logistico_final))
  ) %>%
  modify_caption("**Table 3. Logistic Regression Table**")
tbl_modelo_logistico_final
```

### Forest Plot

```{r}
plot_data_sorted <- tidy(modelo_logistico_final, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term_label = case_when(
      term == "Age" ~ "Age (per year)",
      term == "SexM" ~ "Sex (Male vs Female)",
      term == "ChestPainTypeATA" ~ "Atypical Angina (vs Asympt.)", 
      term == "ChestPainTypeNAP" ~ "Non-Anginal Pain (vs Asympt.)",
      term == "ChestPainTypeTA"  ~ "Typical Angina (vs Asympt.)",
      term == "RestingBP" ~ "Resting BP (mmHg)",
      term == "Cholesterol" ~ "Cholesterol (mg/dL)",
      term == "FastingBS" ~ "Fasting BS (>120 vs Normal)", 
      term == "RestingECGNormal" ~ "ECG Normal (vs LVH)",
      term == "RestingECGST"     ~ "ECG ST-T Abnormality (vs LVH)",
      term == "MaxHR" ~ "MaxHR (per bpm)",
      term == "ExerciseAnginaY" ~ "Exercise Angina (Yes vs No)",
      term == "Oldpeak" ~ "Oldpeak (ST depression)",
      term == "ST_SlopeDown" ~ "ST-Slope Down (vs Flat)",
      term == "ST_SlopeUp"   ~ "ST-Slope Up (vs Flat)",
      TRUE ~ term 
    )
  )
forest_plot_sorted <- ggplot(plot_data_sorted, aes(x = estimate, y = reorder(term_label, estimate))) +

  geom_vline(xintercept = 1.0, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),color = "black") +

  geom_point(size = 3, color = "black") +
  scale_x_log10(breaks = c(0.1, 0.5, 1.0, 2, 5, 10)) + 
  
  labs(
    title = "Forest Plot",,
    x = "Odds Ratio (Log Scale)",
    y = "" 
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30"),
    axis.text.y = element_text(size = 10, color = "black"), 
    plot.margin = unit(c(1, 1, 1, 1), "cm")
  )
forest_plot_sorted
```
